Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2021-02-02 SIR_manuscript.tex
discussion updated

We incorporated the compartment-specific relative testing weights, $W_S$, $W_I$ and $W_R$, to model random testing and \emph{targeted} testing strategies. Here, in the case of targeted testing and for the simplicity and illustraition purposes, we assumed that infected and recovered individuals are tested at three times the \percap rate of susceptible individuals, i.e., $W_I/W_S=3$ and $W_R=W_I$. Note we have not specified a methodology to assigne particular relative testing weights corresponding to a particular targeted testing scenario. 
This part needs to be developed further in future work. 
Modeling different targeted testing strategies, thus test-specific testing weights, requires prior information of the conditional probabilities of getting tested for people in a given compartment. 
This can be implied when we would like to quantify and compare the effect of different levels of test forcus for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing processes.

{\bf The \percap testing intensity, $\rho$;}
Proposition \ref{prop1}, part \ref{p1:rho}, and also Figure \ref{pan} indicates that increasing testing intensity $\rho$ reduces $\Rnum$. This is sensible since as people are moved to test compartments, namely $Z_p$ and $Z_n$ for $Z \in \{S,I,R\}$, they may partially or fully self-isolate. Also individuals who have tested positive, namely $Z_c$ compartment for $Z \in \{S,I,R\}$, are highly likely to self-isolate. The higher probability of being subject to isolation, the lower the force of infection \eqref{Lambda} and the lower $\Rnum$ will become. While our simulation supported this result, analytically it appeared to be hard to conclude this due to the complexity of $\Rnum$ expression.

{\bf Delay in test reporting rate, $\omega$};
Proposition \ref{prop1} part \ref{p1:omega} is the most interesting result. It states that returning test results more rapidly, i.e., increasing $\omega$, does not necessarily lower $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters (test reporting rate, testing processes/strategies ($W_*$), testing efficacy ($P_I$) and isolation ($\eta_*$). Specifically, the $\partial{R_0}/\partial{\omega}>0$ suggests that reproduction number may increase as the test reporting process become faster. This can be seen from expression \eqref{Rom} in the case of perfect isolation, when $\eta_w=0$ and $\eta_c=0$. The underlying mechanism for this result is that individuals may take more precautionary behaviors (e.g., physical distancing) when they are awaiting test results compared to when they are untested or their test has returned negative. An example of a situation that would favour slow return of tests is if the test being employed produces many false negatives, because many infected individuals will believe they are negative and thus may unknowingly spread the virus to many others.  For an example of the behaviour of $\Rnum$ see Figure \ref{pan} at varying values of $\omega$. This property also provides a tool to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$ and consequently control the epidemic.

To give a biological interpretation, we describe the qualitative trends predicted by the inequality. Returning test results more rapidly tends to be more favorable, i.e., reduces $\Rnum$, when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_c \ll \eta_w$), (ii) the test is highly sensitive, i.e., $p_I$ is close to 1, and (iii) the targeted testing strategy is practiced. An in-depth discussion of these results is presented in the discussion section.

- The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID-19 pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. Using our Taylor approximation of $\Rnum$ near $\rho = 0$, we examined what circumstances (i.e., model parameters) make the use of one test more favourable than another, and give a complete description through inequality \ref{eq:rho1vsrho2}. In general, we found that the expensive test tends to more effectively lower $\Rnum$ when (a) individuals who test positive self-isolate much more than individuals who are waiting for their test result, (b) the time it takes to return tests is much shorter than the mean infectious period, and (c) the testing intensity is much greater for infected individuals than susceptible individuals.
- Our model also predicts an asymmetry between cheap and expensive tests: if a cheap test can identify on average more infected individuals as an expensive test, then our model predicts that the cheap test will lower $\Rnum$ more. In contrast, if an expensive test can identify on average more infected individuals, it will not necessarily lower $\Rnum$ more than the cheap test.

{\bf Other points};
one point we can make when discussing the relative weight of testing in different compartments is to discuss pre-testing screening tools, such as surveys or questionnaires. If we have a quantitative description of how the testing intensities affect the dynamics, we can make statements like ``our results suggest that employing a pre-testing screening tool can help target infected individuals more effectively. In particular, doubling the sensitivity of the pre-screening tool would *do something* to $\Rnum$.

@@

\begin{align}\label{eq:necsuf}
    &\omega > \omega_2 \nonumber \\
    &\omega > \frac{\gamma}{\sqrt{1-Q}-1} \nonumber \\
    &\vdots \nonumber \\
    &\frac{1-n_{t}}{1-n_{w}}P_{i}>\frac{W_{s}}{W_{i}}\left(\frac{\gamma}{\omega}+1\right)^{2}
\end{align}

>>>>

We incorporated the compartment-specific relative testing weights, $W_S$, $W_I$ and $W_R$, to model random testing and \emph{targeted} testing strategies. Here, in the case of targeted testing and for the simplicity and illustraition purposes, we assumed that infected and recovered individuals are tested at three times the \percap rate of susceptible individuals, thus $W_I/W_S=3$ and $W_R=W_I$. Note that we have not specified a methodology to assigne particular relative testing weights corresponding to a particular targeted testing scenario. 
This part needs to be developed further in future work. 
Modeling different targeted testing strategies, equivalently test-specific testing weights in our framework, requires prior information of the conditional probabilities of getting tested for people in a given compartment. 
This can be implied when we would like to quantify and compare the effect of different levels of test forcus for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing processes.

{\bf The \percap testing intensity, $\rho$;}
Proposition \ref{prop1} part \ref{p1:rho} and also Figure \ref{pan} indicates that increasing testing intensity $\rho$ reduces $\Rnum$. This is sensible since as people are moved to test compartments, namely $Z_p$ and $Z_n$ for $Z \in \{S,I,R\}$, they may partially or fully self-isolate. Also individuals who have tested positive, namely $Z_c$ compartment for $Z \in \{S,I,R\}$, are highly likely to self-isolate. The higher probability of being subject to isolation, the lower the force of infection \eqref{Lambda} and the lower $\Rnum$ will become. While our simulation supported this result, analytically it appeared to be hard to conclude this due to the complexity of $\Rnum$ expression.

{\bf Delay in test reporting rate, $\omega$;}
Proposition \ref{prop1} part \ref{p1:omega} is the most interesting result. It states that returning test results more rapidly, i.e., increasing $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters (test reporting rate, testing strategies or test-specific testing weights, test sensitivity and specificity and isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta_w=0$ and $\eta_c=0$, $\Rnum$ may increase as the test reporting process become faster. This can be seen from expression \eqref{Rom}.
The interpretation for this phenomenon is that individuals are highly likely to fully self-isolate when they are awaiting test results and when they are reported positive with a negligable influence on the force of infection, $\lambda$. Thus the faster the test reporting rate, the shorter they stay in these ``\emph {safe}'' awaiting compartments, and the more they get involved in the infection process. 
Another example of this situation, when the slow test reporting rate would be favorable is in the case when the test being employed produces many false negatives. Because many infected individuals will believe they are uninfectious and thus may unknowingly spread the virus to many others.

Our model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf} \todo{Fady please confirm}. To summarise, returning test results more rapidly tends to be more favorable, i.e., reduces $\Rnum$, when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_c \ll \eta_w$), (ii) the test is highly sensitive, i.e., $p_I$ is close to 1, and (iii) the targeted testing strategy is practiced.

Our model also predicts an asymmetry between cheap and expensive tests: if a cheap test can identify on average more infected individuals as an expensive test, then our model predicts that the cheap test will lower $\Rnum$ more. In contrast, if an expensive test can identify on average more infected individuals, it will not necessarily lower $\Rnum$ more than the cheap test. The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID-19 pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. Using our Taylor approximation of $\Rnum$ near $\rho = 0$, we examined what circumstances (i.e., model parameters) make the use of one test more favourable than another, and give a complete description through inequality \ref{eq:rho1vsrho2}. In general, we found that the expensive test tends to more effectively lower $\Rnum$ when (a) individuals who test positive self-isolate much more than individuals who are waiting for their test result, (b) the time it takes to return tests is much shorter than the mean infectious period, and (c) the testing intensity is much greater for infected individuals than susceptible individuals.

{\bf Other points [?];}
one point we can make when discussing the relative weight of testing in different compartments is to discuss pre-testing screening tools, such as surveys or questionnaires. If we have a quantitative description of how the testing intensities affect the dynamics, we can make statements like ``our results suggest that employing a pre-testing screening tool can help target infected individuals more effectively. In particular, doubling the sensitivity of the pre-screening tool would *do something* to $\Rnum$.

@@

\todo{Fady, can you clean this up}
\begin{align}\label{eq:necsuf}
    &\omega > \omega_2 \nonumber \\
    &\omega > \frac{\gamma}{\sqrt{1-Q}-1} \nonumber \\
    &\vdots \nonumber \\ 
    &\frac{1-n_{t}}{1-n_{w}}P_{i}>\frac{W_{s}}{W_{i}}\left(\frac{\gamma}{\omega}+1\right)^{2}
\end{align}
